---
title: 'R Tutorial 8: Propensity Score Matching'
date: today
engine: knitr
---

Other ressources: 

- https://www.andrewheiss.com/blog/2021/12/18/bayesian-propensity-scores-weights/ (not so much for the bayesian part)

- https://www.r-causal.org/

- https://cran.r-project.org/web/packages/MatchIt/vignettes/MatchIt.html

(a bunch of vignette that I do not have time to read, seems good)

Idea: students that attends chatolic school are different on average, hence if we want to take into account the 'causal estimates of Catholic schooling' we need to take that into account.

We are trying to see if catholic school have causal effect on "student achievement".

"catholic school" ---> "student achievement"

Here being "catholic school" is considered "Treated" and public school is "Control"

```{r}
#| label: load package
library(MatchIt) # needed to install this one
library(dplyr)
library(ggplot2)
```

```{r}
#| label: load data 
ecls <- read.csv("data-processed/ecls.csv")
```

# Pre-analysis using non-matched data

## Difference-in-means: outcome variable

"student achievement" = **standardized** math score

independent variable of interest: catholic (1 = student went to catholic school; 0 = student went to public school).

```{r}
ecls |>  
  summarise(n_students = n(),
            mean_math = mean(c5r2mtsc_std),
            std_error = sd(c5r2mtsc_std) / sqrt(n_students),
            .by = catholic)
```

t-test

```{r}
with(ecls, t.test(c5r2mtsc_std ~ catholic))
```

## Difference-in-means: pre-treatment covariates

5 covariates:

- race_white: Is the student white (1) or not (0)?

- p5hmage: Mother’s age

- w3income: Family income

- p5numpla: Number of places the student has lived for at least 4 months

- w3momed_hsb: Is the mother’s education level high-school or below (1) or some college or more (0)?

```{r}
ecls_cov <- c('race_white', 'p5hmage', 'w3income', 'p5numpla', 'w3momed_hsb')
ecls |> 
  group_by(catholic) |> 
  select(one_of(ecls_cov))|> 
  summarise_all(funs(mean(., na.rm = T))) # deprecated 
```

Folks in catholic school tend to have more income, more white, slightly stay in one place and the mother is more educated.

#  Propensity score estimation

```{r}
# why do we divide by 1K, should we standardize instead?
ecls <- ecls |> mutate(w3income_1k = w3income / 1000)
# I prefer drawing the confounder first, etc first and even start 
# with a more simple model, It seems this model has confounder  
# see do-calculus
# we can use dagitty 
m_ps <- glm(catholic ~ race_white + w3income_1k + p5hmage + p5numpla + w3momed_hsb,
            family = binomial(), data = ecls)
summary(m_ps)
```

Quite easy:

```{r}
prs_df <- data.frame(pr_score = predict(m_ps, type = "response"),
                     catholic = m_ps$model$catholic)
head(prs_df)
```

## Examining the region of common support

```{r}
labs <- paste("Actual school type attended:", c("Catholic", "Public"))
prs_df |>
  mutate(catholic = ifelse(catholic == 1, labs[1], labs[2])) |>
  ggplot(aes(x = pr_score)) +
  geom_histogram(color = "white") +
  facet_wrap(~catholic) +
  xlab("Probability of going to Catholic school") +
  theme_bw()
```

Is it supposed to be good?

# Executing a matching algorithm

```{r}
ecls_nomiss <- ecls %>%  # MatchIt does not allow missing values
  select(c5r2mtsc_std, catholic, one_of(ecls_cov)) %>%
  na.omit()

mod_match <- matchit(catholic ~ race_white + w3income + p5hmage + p5numpla + w3momed_hsb,
                     method = "nearest", data = ecls_nomiss)
```

```{r}
summary(mod_match)
plot(mod_match)
```


```{r}
dta_m <- match.data(mod_match)
dim(dta_m)
head(dta_m)
```

(I have 10 columns)

# Examining covariate balance in the matched sample

1. visual inspection: it seems" "If matching is done well, the treatment and control groups will have (near) identical means of each covariate at each value of the propensity score." 

2. Difference in means

```{r}
dta_m |>
  group_by(catholic) |>
  select(one_of(ecls_cov)) |>
  summarise_all(funs(mean)) # still deprecated
```

3. Average absolute standardized difference (i did not try but get it)

# Estimating treatment effects

Use dta_ma

```{r}
lm_treat1 <- lm(c5r2mtsc_std ~ catholic, data = dta_m)
summary(lm_treat1)
```
